A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-13 paper/paper.tex
Add more text to the extension section

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion}, a scheme in which logical
relation for dependent types is typically defined~\citep{todo}, we opt
for a direct explanation in informal language of sets, where
impredicativity manifests in the form of second-order logical
formulas and thus more intuitive to understand for readers with some
general mathematics background.

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation. In Section~\ref{sec:discuss}, we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we show that extensions such
as $\eta$ rules for functions and normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}.
Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown or simpler and therefore omitted from the text.

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms.

The updated logical relation and its auxiliary definitions are shown
in Figure~\ref{fig:logrelopen}. When extend parallel reduction to
include $\eta$ reduction for functions. This effectively adds $\eta$
laws for functions to our equational theory since coherence, the
untyped relation used for type conversion, is built on top of parallel
reduction.

We can recover the same confluence result about parallel reduction
using the standard techniques from \citet{barendregt, takahashi},
though specifically for the de Bruijn representation, we need the
following anti-renaming lemma about parallel reduction:
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming}
  If $[[a {xi} => b0]]$, then there exists some $[[b]]$ such that
  $[[b {xi} = b0]]$ and $[[a => b]]$.
\end{lemma}

% The idea of a term blocked from $\beta$ reduction is
% captured by $[[ne]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.
The syntactic forms $[[ne]]$ and $[[nf]]$ capture the neutral terms
and normal forms with respect to $\beta$-reduction, but can still take
$\eta$ reduction steps. Furthermore, we can show that parallel
preserves $[[ne]]$ and $[[nf]]$.

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction.

@@

\begin{lemma}[$CR$ for interpreted sets]
  \label{lemma:crel}
  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The base
  cases are all immediate.

@@

\begin{lemma}[$CR_2$ for type]
  \label{lemma:cr2ty}
  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The
  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
\end{proof}

@@

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.

The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemmas\ref{lemma:cr2tyrec} and
\ref{lemma:crelrec}.

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{theorem}[Existence of $\beta$-normal form]
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{theorem}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]]$ as the term constructor that injects
  natural numbers as variables. This is
  allowed since Lemma~\ref{lemma:crelrec} says neutral terms
  semantically inhabit any type.
\end{proof}

\section{Mechanization}
\label{sec:logrelmech}

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of the \texttt{tstar} function, which stands for the Takahashi
translation~\citep{takahashi-parallel-reduction}. \texttt{inv:Par} and
\texttt{ctrs:Par} says that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} says that the proof would
require the reflexivity and congruence properties of parallel
reduction that we have shown earlier.

The automation here not only gives us a proof that is shorter and more
resiliant to changes in definition, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas that are not specified in the \texttt{use} flag.
\section{Discussion}
\label{sec:discuss}

>>>>

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion}, a scheme in which logical
relation for dependent types is typically defined~\citep{todo}, we opt
for a direct explanation in informal language of sets, where
impredicativity manifests in the form of second-order logical
formulas and thus more intuitive to understand for readers with
general mathematics background.

In Section~\ref{sec:extension}, we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we find that extending the
language with
$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation.

Finally, in Section~\ref{sec:relatedwork}, we discuss how our proof
relates to existing proofs by logical relations for dependent types in
the literature.

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}.
Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown or simpler and therefore omitted from the text.

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
\begin{figure}[h]
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}

In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms. We
first extend parallel reduction to include $\eta$ reduction for
functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
equational theory since coherence, the untyped relation used for type
conversion, is built on top of parallel reduction. We can recover the
same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about
parallel reduction.
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming} If $[[a { xi } => b0]]$, then there
exists some $[[b]]$ such that $[[b { xi } = b0]]$ and $[[a => b]]$.
\end{lemma}

The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
and normal forms with respect to $\beta$-reduction, but can still take
$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
that $[[a = e]]$ or $[[a = f]]$.

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
This lemma is mainly to show that \rref{P-AbsEta} does not introduce
any new $\beta$ redex because $\eta$ reduction is the only real step
we can take.

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.
\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]

The updated logical relation and its auxiliary definitions are shown
in Figure~\ref{fig:logrelopen}.

% The idea of a term blocked from $\beta$ reduction is
% captured by $[[e]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.

@@

\begin{lemma}[$CR$ for interpreted sets]
  \label{lemma:crel}
  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The base
  cases are all immediate.

@@

\begin{lemma}[$CR_2$ for type]
  \label{lemma:cr2ty}
  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The
  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
\end{proof}

@@

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.
The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
\ref{lemma:crelrec}.

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]](i) = i$, which injects
  natural numbers as term variables. This is
  allowed since Lemma~\ref{lemma:crelrec} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho} = A]]$ and $[[a {rho} = a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}
From Corollary~\ref{corollary:exbetanf}, we can show that there exists
$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
and also strictly decreases the size of the term.

Due to the non-deterministic nature of parallel reduction, we need to
take a few more steps to convert the existence of $\beta\eta$-normal
form into a decision procedure for type conversion. More specifically,
we can show that a deterministic evaluation strategy such as
leftmost-outermost reduction can always find the $\beta\eta$-normal
form if there exists one. However, we omit such proofs since they can
can be formulated on untyped lambda terms and thus orthogonal to the
specifics of dependently typed systems. Instead, we redirect readers
to \citet{factorization-essentially, takahashi-parallel-reduction} for
the details.

Finally, we want to point out that
Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
\ref{lemma:crelrec}, often referred to as adequacy of the logical
relation, are in fact required in the normalization proof for simply
typed languages~\citep{abel2019poplmark}. In the simply typed
scenario, adequacy needs to be proven before the fundamental theorem
so we can handle rules such as \rref{T-If} where the scrutinee is a
neutral term. In a simply typed setting, Lemma~\ref{lemma:extwn} is
used in the exact same way to show that lambda terms are themselves
normalizing as we have done in Lemma~\ref{lemma:crel}.

Dependent types make the proof slightly more complicated
as we also need to know that every type has a normal form. Whereas the
$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
requires the $CR_1$ property about the interpreted sets
(Lemma~\ref{lemma:crel}).

Overall, despite the dependently typed setting,
the extension of our logical relation to prove normalization of open
\emph{and} closed terms closely mirrors the progression from
normalization of closed terms~\citep{harpertait} to normalization of
open terms~\citep{harperkripke} in the simply typed lambda calculus.
It is in fact reassuring that once we have laid the foundational
technique for handling dependent types in our logical relation, the
further extensions more or less boil down to properties that can be
derived through syntactic means, which tend to not get in the way of
Gödel's second incompleteness theorem.

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  TODO
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the statics of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
extension from Section~\ref{sec:extension}.

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of the \texttt{tstar} function, which stands for the Takahashi
translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
\texttt{ctrs:Par} says that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
automation tactic to use
the reflexivity and congruence properties of parallel
reduction as lemmas.

The automation here not only gives us a proof that is shorter and more
resiliant to changes in definition, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas or invert derivations that are not specified in the
\texttt{use} or \texttt{inv} flag.
\section{Discussion}
\label{sec:discuss}
